---
title: "TEROS54 Data Processing Tutorial"
author: "Moreen Willaredt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TEROS54 Data Processing Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates how to use the `teros54reader` package to process and visualize data from METER Group TEROS54 soil moisture and temperature sensors. The TEROS54 is a popular multi-depth soil sensor that measures volumetric water content and temperature at up to 4 depths per sensor.

## Package Overview

The `teros54reader` package provides:

- **Data processing**: Read and parse TEROS54 Excel files
- **Data visualization**: Multiple plotting functions for different analysis needs
- **Quality control**: Built-in data validation and cleaning

## Loading the Package

```{r}
library(teros54reader)
```

## Reading TEROS54 Data

The main function `process_sensor_data()` reads Excel files from TEROS54 dataloggers and converts them into a clean, analysis-ready format.

```{r}
# Get path to sample data included with package
sample_file <- system.file("extdata", "sample_teros54_data.xlsx", package = "teros54reader")

# Process the data
soil_data <- process_sensor_data(sample_file, verbose = TRUE)

# View the structure
str(soil_data)
```

```{r}
# First few rows of data
head(soil_data)
```

The processed data contains:
- `time`: Timestamp of measurements
- `wc`: Volumetric water content (m³/m³)
- `temp`: Soil temperature (°C)
- `depth`: Measurement depth (cm)
- `sensor_id`: Sensor identifier (Port 1, Port 2, etc.)
- `wc_valid`, `temp_valid`: Data quality flags

## Basic Data Exploration

```{r}
# Summary of the data
summary(soil_data)
```

```{r}
# Number of sensors and measurements
cat("Number of sensors:", length(unique(soil_data$sensor_id)), "\n")
cat("Number of measurements per sensor:", table(soil_data$sensor_id), "\n")
cat("Measurement depths:", levels(soil_data$depth), "cm\n")
cat("Time range:", format(range(soil_data$time), "%Y-%m-%d %H:%M"), "\n")
```

## Water Content Visualization

### Time Series Plots

```{r}
# Plot water content time series for all sensors
plot_water_content(soil_data, title = "Sample Site Water Content")
```

```{r}
# Plot specific sensors with custom layout
plot_water_content(soil_data, 
                  sensors = c("Port 1", "Port 2"), 
                  ncol = 2,
                  title = "Comparison of Port 1 and Port 2")
```

### Soil Profile Plots

```{r}
# Water content profiles at recent time points
plot_soil_profile(soil_data, plot_var = "wc")
```

## Temperature Visualization

```{r}
# Plot temperature time series
plot_temperature(soil_data, title = "Sample Site Temperature")
```

```{r}
# Temperature profiles
plot_soil_profile(soil_data, plot_var = "temp")
```

## Sensor Comparison

```{r}
# Compare all sensors side by side
compare_sensors(soil_data, plot_var = "wc")
```

```{r}
# Compare only surface measurements
compare_sensors(soil_data, plot_var = "temp", depths = c("15", "30"))
```

## Quick Plotting

For rapid data exploration, use the `quick_plot_teros54()` function:

```{r eval=FALSE}
# Plot directly from file (both water content and temperature)
plots <- quick_plot_teros54(sample_file)

# Display water content plot
plots$water_content

# Display temperature plot  
plots$temperature
```

```{r}
# Just water content
quick_plot_teros54(sample_file, plot_type = "water")
```

## Data Quality Assessment

The package includes data quality flags to help identify potentially problematic measurements:

```{r}
# Check data quality
cat("Water content data quality:\n")
table(soil_data$wc_valid)

cat("\nTemperature data quality:\n")
table(soil_data$temp_valid)
```

```{r}
# Filter to only valid measurements
clean_data <- soil_data[soil_data$wc_valid & soil_data$temp_valid, ]
cat("Rows after quality filtering:", nrow(clean_data), "of", nrow(soil_data), "\n")
```

## Advanced Analysis Examples

### Daily Patterns

```{r}
# Add hour of day for pattern analysis
soil_data$hour <- as.numeric(format(soil_data$time, "%H"))

# Plot daily patterns for surface sensors (15cm depth)
library(ggplot2)
surface_data <- soil_data[soil_data$depth == "15", ]

ggplot(surface_data, aes(x = hour, y = wc, color = sensor_id)) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "Daily Water Content Patterns (15cm depth)",
       x = "Hour of Day",
       y = "Water Content (m³/m³)",
       color = "Sensor") +
  theme_minimal()
```

### Depth Gradients

```{r}
# Calculate mean values by depth and sensor
library(dplyr)
depth_summary <- soil_data %>%
  group_by(sensor_id, depth) %>%
  summarise(
    mean_wc = mean(wc, na.rm = TRUE),
    mean_temp = mean(temp, na.rm = TRUE),
    .groups = "drop"
  )

print(depth_summary)
```

## Tips for Real Data

When working with your own TEROS54 data:

1. **File format**: Ensure your Excel files have the standard TEROS54 logger format
2. **Timezone**: Use the `timezone` parameter if your data isn't in UTC
3. **Logger data**: Set `include_logger = TRUE` if you need battery/barometer data
4. **Quality control**: Always check the `*_valid` flags before analysis
5. **Memory**: For large datasets, consider processing in chunks or filtering by date range

## Conclusion

The `teros54reader` package streamlines TEROS54 data processing and visualization, making it easy to go from raw logger files to publication-ready plots. The package handles the complex Excel parsing automatically and provides flexible visualization options for different analysis needs.

For more information and updates, visit: https://github.com/UrbanLab-UCD/teros54-reader